---
title: "Quickstart for Model Documentation (R)"
author: "ValidMind"
date: "2026-03-18"
output: html_document
---

# Quickstart for Model Documentation

Learn the basics of using ValidMind to document models as part of a model development workflow using R. This notebook uses the ValidMind R package (a `reticulate` wrapper around the Python library) to generate a draft of documentation for a binary classification model.

We will:

1. Import a sample dataset and preprocess it
2. Split the datasets and initialize them for use with ValidMind
3. Train a logistic regression (GLM) model and initialize it for use with testing
4. Run the full suite of documentation tests, sending results to the ValidMind Platform

## Setting up

Set `python_version` to the path of the Python binary that has `validmind` installed.
To find it, run `which python` in your terminal with the environment activated.

```{r setup, include=FALSE}
library(reticulate)

# Replace with the path to your Python binary, e.g.:
# python_version <- "/Users/you/project/.venv/bin/python"
python_version <- "<path_to_python_version>"
use_python(python_version)

library(validmind)
library(dplyr)
library(caTools)
library(knitr)

knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

## Initialize the ValidMind Library

Log in to the [ValidMind Platform](https://app.prod.validmind.ai) and register a model:

1. Navigate to **Inventory** and click **+ Register Model**.
2. Under **Documents > Development**, select the `Binary classification` template.
3. Go to **Getting Started**, select `Development` from the **DOCUMENT** drop-down, and copy the code snippet.

Replace the placeholder values below with your own credentials:

```{r}
vm_r <- vm(
  api_host = "https://app.prod.validmind.ai/api/v1/tracking",
  api_key = "<your_api_key_here>",
  api_secret = "<your_api_secret_here>",
  model = "<your_model_id_here>",
  python_version = python_version,
  document = "documentation"
)
```

## Preview the documentation template

Verify the connection and see the documentation structure:

```{r}
vm_r$preview_template()
```

## Load the demo dataset

We use the Bank Customer Churn dataset for this demonstration:

```{r}
data <- read.csv("notebooks/code_sharing/datasets/bank_customer_churn.csv")
head(data)
```

## Initialize the raw dataset

Before running tests, initialize a ValidMind dataset object for the raw data:

```{r}
vm_raw_dataset <- vm_r$init_dataset(
  dataset = data,
  input_id = "raw_dataset",
  target_column = "Exited",
  class_labels = list("0" = "Did not exit", "1" = "Exited")
)
```

## Preprocess the raw dataset

Handle categorical variables using one-hot encoding and remove unnecessary columns:

```{r}
data_processed <- data %>% select(-RowNumber, -CustomerId, -Surname)

# One-hot encode categorical variables
geo_dummies <- model.matrix(~ Geography - 1, data = data_processed)
gender_dummies <- model.matrix(~ Gender - 1, data = data_processed)
data_processed <- data_processed %>% select(-Geography, -Gender)
data_processed <- cbind(data_processed, geo_dummies, gender_dummies)
```

### Split the dataset

Split into training (60%), validation (20%), and test (20%) sets:

```{r}
set.seed(42)

# First split: 80% train+validation, 20% test
split1 <- sample.split(data_processed$Exited, SplitRatio = 0.8)
train_val_data <- subset(data_processed, split1 == TRUE)
test_data <- subset(data_processed, split1 == FALSE)

# Second split: 75% train, 25% validation (of the 80% = 60/20 overall)
split2 <- sample.split(train_val_data$Exited, SplitRatio = 0.75)
train_data <- subset(train_val_data, split2 == TRUE)
validation_data <- subset(train_val_data, split2 == FALSE)
```

## Train a logistic regression model

Train a GLM with a binomial family (logistic regression):

```{r}
model <- glm(Exited ~ ., data = train_data, family = binomial)
summary(model)
```

## Initialize the ValidMind datasets

```{r}
vm_train_ds <- vm_r$init_dataset(
  dataset = train_data,
  input_id = "train_dataset",
  target_column = "Exited"
)

vm_test_ds <- vm_r$init_dataset(
  dataset = test_data,
  input_id = "test_dataset",
  target_column = "Exited"
)
```

## Initialize a model object

Save the R model and initialize it with ValidMind:

```{r}
model_path <- save_model(model)

vm_model <- vm_r$init_r_model(
  model_path = model_path,
  input_id = "model"
)
```

### Assign predictions

Link model predictions to the training and testing datasets:

```{r}
vm_train_ds$assign_predictions(model = vm_model)
vm_test_ds$assign_predictions(model = vm_model)
```

## Run the full suite of tests

Build the test configuration that maps each test to its required inputs:

```{r}
# Import the test config helper from the Python customer_churn module
customer_churn <- reticulate::import("validmind.datasets.classification.customer_churn")
test_config <- customer_churn$get_demo_test_config()
```

Preview the test configuration:

```{r}
vm_utils <- reticulate::import("validmind.utils")
vm_utils$preview_test_config(test_config)
```

Run the full documentation test suite and upload results to the ValidMind Platform:

```{r}
full_suite <- vm_r$run_documentation_tests(config = test_config)
```

## Next steps

Head to the [ValidMind Platform](https://app.prod.validmind.ai) to view the generated documentation:

1. Navigate to **Inventory** and select your model.
2. Click **Development** under Documents to see the full draft of your model documentation.

From there, you can make qualitative edits, collaborate with validators, and submit for approval.

---

*Copyright 2023-2026 ValidMind Inc. All rights reserved.*
*Refer to [LICENSE](https://github.com/validmind/validmind-library/blob/main/LICENSE) for details.*
*SPDX-License-Identifier: AGPL-3.0 AND ValidMind Commercial*
